---
title: "Fig. 3"
output:
  word_document: default
  html_notebook: default
---

```{r}
library(cowplot)
library(dplyr)
library(ggplot2)
library(ggmap)
library(devtools)
install_github("oswaldosantos/ggsn")
library(ggsn)
library(ggthemes)
library(raster)
library(readxl)
library(reshape2)
library(rgdal)
library(sp)
library(tidyr)
library(viridis)
library(xlsx)
```

## Set up geodata

```{r}
epsg <- 4326
crs <- paste("+init=epsg:",epsg,"", sep="")

cnt <- spTransform(readOGR(dsn = "data/vector", layer="10m_admin_0_countries", verbose = FALSE), CRS( crs))
rvr <- spTransform(readOGR(dsn = "data/vector", layer="10m_rivers_lake_centerlines", verbose = FALSE), CRS(crs))

# raster layer with rainforest cover from bioval.jrc.ec.europa.eu
dpath <- "data/raster/bioval.jrc.ec.europa.eu/hdr.adf"
x <- new("GDALReadOnlyDataset", dpath)
getDriver(x)
getDriverLongName(getDriver(x))

hdr <- asSGDF_GROD(x)
hdr <- raster(hdr)

# extent format (xmin,xmax,ymin,ymax)
e  <- extent(6, 23, -8.5, 8.5) 
rfs <- crop(hdr, e) 

# convert the raster to points for plotting
rfs.p <- rasterToPoints(rfs)

# Make the points a dataframe for ggplot & subset rainforest bands 1-7
rfs.d <- data.frame(rfs.p)
rfs.d <- subset(rfs.d, band1 >= 1 & band1 <= 7)

# Create basemap
g.b <- ggplot() + 
  geom_polygon(data = cnt, 
              aes(long, lat, group = group, fill=hole),
              fill  = "#ffebbe") + 
  geom_raster(data = rfs.d, aes(y = y, x = x), fill = '#00734d') + 
  geom_polygon(data = cnt, 
              aes(long, lat, group = group),
              size = 0.2, 
              fill  = NA,  
              color = "black") + 
  geom_path(data = rvr, 
            aes(long, lat, group = group, fill=NULL),
            size = 0.5,   
            colour  = "#44afe3") + 
  scalebar(location = "bottomright", dist = 250, height = 0.2,
           dd2km = TRUE, model = 'WGS84',           
           x.min = 15, x.max = 22,
           y.min = -6.25, y.max = -5.25, 
           st.dist = .4, st.size = 4) +
  coord_equal(xlim=c(7, 22),
              ylim=c(-7.5,7.5)) + 
  theme_few() + 
  theme(
    panel.background = element_rect(fill = "#dff1f9")
  )

# the minimap:
csl <- spTransform(readOGR(dsn = "data/vector", layer="ne_110m_land", verbose = FALSE), CRS( crs))

g2 <- ggplotGrob(
  ggplot() + 
  geom_path(data = csl, 
               aes(long, lat, group = group),
               size = 0.2, 
               fill = 'white', color = "black") + 
  geom_rect(mapping = aes(xmin = 7, xmax = 22, 
                          ymin = -7.5, ymax = 7.5), 
            alpha = 0.5, fill ="black", size = 0.5) + 
  coord_equal() + 
  scale_x_continuous(limits = c(-20, 55), expand = c(0, 0)) + 
  scale_y_continuous(limits = c(-40, 42), expand = c(0, 0)) + 
  theme_map() + 
  theme(panel.background = element_rect(fill = 'white'), 
        panel.border = element_rect(colour  = 'black', fill = NA, size = 1))
)
```

# Database S1. Compiled database of 14C-dated archaeological sites in Western Central Africa

## read dataset & filter only dates between 3000-2000 bp

```{r}
s <- read_xlsx("data/base/pnas.1715336115.sd01.xlsx", 
                skip = 1) %>%
  filter(`Date 14C BP` >= 2000 & `Date 14C BP` <= 3000)
s
```

## review dataset

### clear misspelled sites

```{r}
s[s[,13] == "Abang minko'o", 13] <- "Abang Minko'o"
s[s[,13] == 'Boso Njafo', 13] <- 'Boso-Njafo'
s[s[,13] == 'Mintyaminyoumin, MNT 08/1', 13] <- 'Mintyaminyoumin'
s[s[,13] == 'Mintyaminyoumin, MNT 08/2', 13] <- 'Mintyaminyoumin'
s[s[,13] == 'Issemeyo 4-2', 13] <- 'Issemeyo'
s[s[,13] == 'Issemeyo 4-19', 13] <- 'Issemeyo'
s[s[,13] == 'Issemeyo 7-1', 13] <- 'Issemeyo'
s[s[,13] == 'Mont Brazza Okanda 1', 13] <- 'Okanda 1'
s[s[,13] == 'Bwambé', 13] <- 'Bwambé-Sommet'
s[s[,13] == 'Bwambé-Sommet, BWS 04/1', 13] <- 'Bwambé-Sommet'
s[s[,13] == 'Bwambé-Sommet, BWS 04/2', 13] <- 'Bwambé-Sommet'
s[s[,13] == 'Bwambé-Sommet, BWS 04/3', 13] <- 'Bwambé-Sommet'
s[s[,13] == 'Campo 1', 13] <- 'Campo'
s[s[,13] == 'Campo, CAM 07/11', 13] <- 'Campo'
s[s[,13] == 'Ebel Alembe 1-1', 13] <- 'Ebel Alembe'
s[s[,13] == 'Ebel Alembe 1-2', 13] <- 'Ebel Alembe'
s[s[,13] == 'Ebel Alembe 1-4', 13] <- 'Ebel Alembe'
s[s[,13] == 'Ebel Alembe 2-2', 13] <- 'Ebel Alembe'
```

### change wrong coordiantes
```{r}
# Boso-Njafo (see Kahlheber et al. 2014)
s[s[,13] == 'Boso-Njafo', 2] <- 1.130547
s[s[,13] == 'Boso-Njafo', 3] <- 19.209599

# Gba
s[s[,13] == 'Gba', 2] <- 5.65
s[s[,13] == 'Gba', 3] <- 11.459

# Ikengué 1
s[s[,13] == 'Ikengué 1', 2] <- -1.5
s[s[,13] == 'Ikengué 1', 3] <- 9.4986

# Kango 5
s[s[,13] == 'Kango 5', 2] <- 0.200800
s[s[,13] == 'Kango 5', 3] <- 10.0825

# Moanda 1
s[s[,13] == 'Moanda 1', 2] <- -1.556
s[s[,13] == 'Moanda 1', 3] <- 13.181

# Moanda 2
s[s[,13] == 'Moanda 2', 2] <- -1.5556
s[s[,13] == 'Moanda 2', 3] <- 13.1806

# Mvindou
s[s[,13] == 'Mvindou', 2] <- -4.472
s[s[,13] == 'Mvindou', 3] <- 11.952

# Ngoume PK 35
s[s[,13] == 'Ngoume PK 35', 2] <- 5.483
s[s[,13] == 'Ngoume PK 35', 3] <- 11.4

# Obobogo
s[s[,13] == 'Obobogo', 2] <- 3.817
s[s[,13] == 'Obobogo', 3] <- 11.504

# Okolo
s[s[,13] == 'Okolo', 2] <- 3.9333
s[s[,13] == 'Okolo', 3] <- 11.5167

# Pikounda 1 - should be 0.549344	16.635736
s[s[,13] == 'Pikounda 1', 2] <- 0.549
s[s[,13] == 'Pikounda 1', 3] <- 16.636

# Epona 2
s[s[,13] == 'Epona 2', 2] <- -0.106
s[s[,13] == 'Epona 2', 3] <- 11.147

# Ngoume PK 35
s[s[,13] == 'Ngoume PK 35', 2] <- 5.483
s[s[,13] == 'Ngoume PK 35', 3] <- 11.4
```

### remove dates from the ICB that HPW considers 'not-represenative' (see Seidensticker 2017: Tab. 21)
```{r}
s <- s[!s[,1] == "Hv-11570", ] # Bamanya 12 BAM 81/1 Hv-11570 2245 ± 195 bp - - ebd., 412
s <- s[!s[,1] == "Hv-11575", ] # Imbonga 43 IMB 81/1 Hv-11575 2130 ± 125 bp Monkoto - ebd., 99
s <- s[!s[,1] == "Hv-12611", ] # Wafanya 58 WAF 83/16  2695 ± 160 bp Monkoto/Longa - ebd., 99 Tab. 32, 127 Tab. 53
s <- s[!s[,1] == "Hv-12614", ] # Imbonga 43 IMB 83/1 Hv-12614 2665 ± 110 bp Monkoto - ebd., 99 Tab. 32
s <- s[!s[,1] == "Hv-12615", ] # Bamanya 12 BAM 83/2 Hv-12615 2210 ± 180 bp - - ebd., 412
```

### remove sites without propper context details (BC)
```{r}
rem <- c("Abala", 
"Area 1", 
"Bafounda", 
"Bai Dédé Creek 1",
"Boundji",
"Fiye Nkwi",
"Gangalingolo",
"Goalake river 2",
"Gombe",
"Ikelemba 1",
"Ilamba 2",
"Kenie Branch 3",
"Kingabwa",
"KS8",
"Kwimba",
"Locality 24-5 (Bertoua)",
"Lukula",
"Mbi Crater",
"Ngoumé Pk 36",
"Oliga",
"Otoumbi 2",
"Plateau de Mpassa",
"Pointe Noire",
"Rivière Denis 2",
"Sablières",
"Sangmelima",
"Shum Laka",
"Site 1 or Area 3, T1",
"Site 1 or Area 3, T3",
"Site 2 or Area 2, T4",
"Site 2 or Area 2, T5",
"Site 2 or Area 2, T5",
"Site 2 or Area 2, T6",
"Upper Esimbi 2",
"Wali 5",
"West Ndoki Lombé 2")

s <- s[!s$`Site name` %in% rem, ]
nrow(s)
```

## from long to wide with oldest date 14C

> or better mean?

```{r}
s$dummy <- 'maxAge'

s.piv <- dcast(s, `Site name` + `Latitude (°)` + `Longitude (°)` ~ dummy, 
               value.var = "Date 14C BP", 
               fun.aggregate = max)
s.piv
write.xlsx(s.piv, "data/processed/pnas.1715336115.sd01_UNIQUE-SITES-3000-2000bp.xlsx")
s.piv
```

## Figures

### Basic map of sites

```{r}
g1 <- g.b + 
  geom_point(data = s.piv, aes(x = `Longitude (°)`, y = `Latitude (°)`), shape = 21, fill = 'black', size = 3, color = "white") +
  geom_point(aes(x = 9.401703, y = 4.661158), shape = 23, size = 3, fill = 'yellow')

g3 <- g1 + 
  annotation_custom(grob = g2, 
                    xmin = 6, xmax = 10,
                    ymin = -8, ymax = -4)
g3
ggsave(g3, filename = "output/Fig3rev_Suppl_Sites.pdf", 
       width = 6, height = 6)
```

### Map with dated sites, age as fill

> %>% arrange(maxAge) changes the order of points and brings the early sites to the 'top'

```{r}
g1 <- g.b + 
  geom_point(data = s.piv %>% arrange(maxAge), 
             aes(x = `Longitude (°)`,
                 y = `Latitude (°)`,
                 fill = cut(maxAge, c(3001, 2800, 2600, 2400, 2200, 1999))), 
             shape = 21, 
             size = 3, 
             color = "white") + 
  scale_fill_viridis(option = "C", 
                     direction = -1, 
                     discrete = T,
                     labels=c("2200-2000 bp",
                              "2400-2200 bp", 
                              "2600-2400 bp", 
                              "2800-2600 bp",
                              "3000-2800 bp", 
                              "")) + 
  geom_point(aes(x = 9.401703, y = 4.661158), shape = 23, size = 3, fill = 'yellow') +
  guides(fill = guide_legend(title="", 
                             reverse = T,
                             ncol = 2)) + 
  theme(legend.position="top")

g3 <- g1 + 
  annotation_custom(grob = g2, 
                    xmin = 6, xmax = 10,
                    ymin = -8, ymax = -4)
g3
ggsave(g3, filename = "output/Fig3rev_Suppl_maxAge-cmap.pdf", 
       width = 6, height = 7)
```

### Count of sites (maxAge) per century

> cut(maxAge, c(3001, 2800, 2600, 2400, 2200, 1999))

```{r}
#ggplot(s, aes(`Date 14C BP`)) + 
#  geom_histogram(binwidth = 100)
```

```{r}
h1 <- ggplot(s.piv, aes(x = (maxAge * -1))) + 
  geom_histogram(breaks = seq(-3000, -2000, by = 100), 
                 col="white", 
                 fill = "#666666") + 
  scale_x_continuous("Age (bp)", 
                     breaks = seq(-3000, -2000, by = 100), 
                     labels = seq(3000, 2000, by = -100)) + 
  theme_bw() + 
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
h1
```

### OxCal

```{r}
library(oxcAAR)
quickSetupOxcal()
#setOxcalExecutablePath("OxCal")
my_dates <- R_Date(s$`Sample code`, s$`Date 14C BP`, s$`Standard error`)
my_sum <- oxcal_Sum(my_dates)

write(my_sum, file = "data/processed/Suppl_TO_OxCal.txt")
```

> run in OxCal an Export "Raw data"

```{r}
sum <- read.table(file = "data/processed/Suppl_FROM_OxCal.prior")

p2 <- ggplot(sum, aes(x = V1, y = V2)) + 
  #annotate("rect", xmin = (1950-2020), xmax = (1950-2600), ymin = 0, ymax = .7, fill = 'orange', alpha = .2) + 
  #annotate("text", x = -1700, y = .45, label = "Lake Barombi\nLHRC") + 
  geom_area(fill = '#666666') + 
  scale_x_continuous("Age (cal BC)", 
                     limits = c(-1000, 0), 
                     breaks = seq(-1000, 0, by = 100), 
                     labels = seq(1000, 0, by = -100)) + 
  scale_y_continuous("Probability density") + 
  theme_bw() + 
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
p2
```

### Build comined figure

```{r}
p <- plot_grid(g3, h1, p2, labels = "AUTO", ncol = 1, align = "v", rel_widths = c(1, 1, 1), rel_heights = c(3, 1, 1))
ggsave("output/Fig3rev_Suppl.pdf", p, width = 5, height = 10)
p
```


# aDRAC

```{r}
d <- read.csv("https://raw.githubusercontent.com/dirkseidensticker/aDRAC/master/data/aDRAC.csv")

# filter rough bounding box
d <- filter(d, LAT < 10 & LAT > -10 & LONG > 5 & LONG < 22)

# clean MATERIAL
# only differntiate between 'Pennisetum glaucum', 'Elaeis guineensis' and others
d$MATERIAL <- as.character(d$MATERIAL)
d$MATERIAL[!d$MATERIAL == "Pennisetum glaucum" & !d$MATERIAL == "Elaeis guineensis"] = "Other material"

# pivot on MATERIAL
d.piv <- dcast(d, LAT + LONG ~ MATERIAL)
d.piv
```


```{r}
p1 <- g.b + 
  geom_point(data = subset(d.piv, d.piv[,'Other material'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'black', size = 3, color = "white") + 
  geom_point(data = subset(d.piv, d.piv[,'Elaeis guineensis'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'orange', size = 3, color = "white") + 
  geom_point(data = subset(d.piv, d.piv[,'Pennisetum glaucum'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'red', size = 3, color = "white") + 
  geom_point(aes(x = 9.401703, y = 4.661158), shape = 23, size = 3, fill = 'yellow')
```

## OxCal

```{r}
library(oxcAAR)
quickSetupOxcal()
#setOxcalExecutablePath("OxCal")
my_dates <- R_Date(d$LABNR, d$C14AGE, d$C14STD)
my_sum <- oxcal_Sum(my_dates)

write(my_sum, file = "data/processed/aDRAC_TO_OxCal.txt")
```

> run in OxCal an Export "Raw data"

```{r}
sum <- read.table(file = "data/processed/aDRAC_FROM_OxCal.prior")

p2 <- ggplot(sum, aes(x = V1, y = V2)) + 
  annotate("rect", xmin = (1950-2020), xmax = (1950-2600), ymin = 0, ymax = .55, fill = 'orange') + 
  annotate("text", x = -1700, y = .45, label = "Lake Barombi\nLHRC") + 
  geom_area() + 
  scale_x_reverse("Age (cal y BC/AD)", limits = c(2000, -10000)) +
  scale_y_continuous("Probability density") + 
  theme_bw() + 
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
```

## Figure


```{r}
p <- plot_grid(p1, p2, labels = "AUTO", ncol = 1, align = "v", rel_widths = c(1, 1), rel_heights = c(3, 1))
ggsave("output/Fig3rev_aDRAC.pdf", p, width = 10, height = 13)
p
```


```{r}
ggplot(sum, aes(x = V1, y = V2)) + 
  annotate("rect", xmin = (-20), xmax = (-600), ymin = .55, ymax = .65, fill = 'orange') + 
  annotate("text", x = ((-20+-600)/2), y = .6, label = "LHRC", color = "white", fontface = 2) + 
  annotate("rect", xmin = (536), xmax = (660), ymin = .55, ymax = .65, fill = 'cornflowerblue') + 
  annotate("text", x = ((536+660)/2), y = .53, label = "LALIA") + 
  annotate("rect", xmin = (700), xmax = (1100), ymin = .55, ymax = .65, fill = 'red') + 
  annotate("text", x = ((700+1100)/2), y = .6, label = "MWP", color = "white", fontface = 2) + 
  annotate("rect", xmin = (1250), xmax = (1850), ymin = .55, ymax = .65, fill = 'blue') + 
  annotate("text", x = ((1250+1850)/2), y = .6, label = "LIA", color = "white", fontface = 2) + 
  geom_area() + 
  scale_x_continuous("v. Chr. / n. Chr.", 
                     limits = c(-4000, 2000),
                     breaks = seq(-4000, 2000, by = 1000), 
                     labels = seq(4000, -2000, by = -1000), 
                     expand=c(0,0)) + 
  scale_y_continuous("Probability density",
                     expand=c(0,0)) + 
  theme_bw() + 
  theme(panel.grid.major.x = element_blank(), 
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank())
ggsave("output/aDRAC-OxCal_4000-0bp.pdf", 
       width = 6, height = 2)
```


## Map only site with dates between 3000-2000 bp (roughly the time of the LHRC)

```{r}
d.r <- filter(d, C14AGE < 3000 & C14AGE > 2000)

# pivot on MATERIAL
d.r.piv <- dcast(d.r, LAT + LONG ~ MATERIAL)
d.r.piv
```

```{r}
g1 <- g.b + 
  geom_point(data = subset(d.r.piv, d.r.piv[,'Other material'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'black', size = 3, color = "white") + 
  geom_point(data = subset(d.r.piv, d.r.piv[,'Elaeis guineensis'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'orange', size = 3, color = "white") + 
  geom_point(data = subset(d.r.piv, d.r.piv[,'Pennisetum glaucum'] != 0), aes(x = LONG, y = LAT), shape = 21, fill = 'red', size = 3, color = "white") + 
  geom_point(aes(x = 9.401703, y = 4.661158), shape = 23, size = 3, fill = 'yellow')

g3 <- g1 + 
  annotation_custom(grob = g2, 
                    xmin = 6, xmax = 10,
                    ymin = -8, ymax = -4)
g3
ggsave(g3, filename = "output/Fig3rev_aDRAC_3000-2000bp.pdf", 
       width = 6, height = 6)
```


